\abstractpage{
Formal theorem proving in Lean 4 remains difficult because proof search receives sparse binary feedback
and must explore a large combinatorial space of tactics, intermediate lemmas, and decomposition choices.
This thesis presents \textbf{GammaZero}, a hierarchical theorem-proving framework that combines local
proof generation with structured skeleton decomposition inside an asymmetric AND/OR search graph.
GammaZero uses scaffold-aware subgoal solving: a child goal is solved inside the Lean proof context in
which it was created, rather than as an independent theorem. The system also records partially useful
failed generations through AST-guided Sorrifier patching and dependency-aware reward computation over
Lean expression trees, allowing proof search to produce structured trajectories rather than only binary
success or failure labels.

Experiments are conducted in Lean 4 with Mathlib pinned to a fixed revision and use
\texttt{Gemini 3 Flash Preview} as the proposal model. On miniF2F-test, GammaZero solves 189 out of
244 theorems at a 512-node cumulative search budget, achieving a placeholder-free pass rate of
\textbf{77.5\%}. The pass rate rises consistently with budget: 59.8\% at 32 nodes, 66.4\% at 64 nodes,
73.8\% at 128 nodes, and 75.8\% at 256 nodes. The 512-node setting uses an average of 30.2K generated
tokens per solved instance under early stopping, and the 128-node setting already reaches 73.8\% with
15.4K tokens per solved instance.

Against recent published miniF2F-test systems reported under Lean 4.17, GammaZero exceeds
Kimina-Prover-Preview-Distill at 1024 samples (70.8\%) and Kimina-Preview + Apollo (75.0\%) while
using a compact cumulative node budget. On Lean 4.29 reruns over the same 244-problem test split,
direct Kimina sampling reaches 53.28\%, Apollo + Kimina reaches 53.69\%, direct Gemini 3 Flash
Preview solves 113/244 theorems (46.3\%), and GammaZero reaches 77.5\%. An internal ablation treats the 32-node result
as the root-only setting: increasing the search
budget to the full 512-node hierarchical setting improves performance from 146/244 solved theorems
(59.8\%) to 189/244 (77.5\%), an absolute gain of 17.7 percentage points. These results support the
central claim that scaffold-aware hierarchical search can turn local proof generation into substantially
higher Lean-verified solve rates while producing scored search data suitable for future policy optimization.
}
